Towards a Verified Tableau Prover for a Quantifier-Free Fragment of Set Theory

### 2023-03-03 paper/CADE/paper.tex
First submitted version of paper

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a quantifier-free fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  Furthermore, we extend the calculus with a light-weight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{Decision procedures \and Tableau \and Proof assistants \and Set theory}
\end{abstract}

\section{Introduction}
In Isabelle/HOL, there are specialised procedures for dealing with e.g.\ natural numbers, linear arithmetic, and metric spaces.
Some of these procedures\todo{Related work?} have been verified in Isabelle/HOL such as a procedure for Presburger arithmetic~\cite{presburger} that was later extended to mixed real-integer arithmetic~\cite{arithmetic}.
This procedure, though, uses reflection to work on goals in Isabelle/HOL, which, during execution, either sacrifices\todo{Formulation: no sacrifices but trust} speed by going through the simplifier or soundness by trusting the code generator.
More recently, \citeauthor{orders}~\cite{orders} presented a verified decision procedure for orders that produces certificates. 
This approach offers efficient execution by using generated code as well as soundness because the certificates are replayed through Isabelle's inference kernel.

@@

Since \MLSS{} admits a tableau calculus, we expect that the generation of certificates\todo{Formulation} will be straightforward.
Like with the aforementioned order solver, this would pave the way towards an integration of the decision procedure into Isabelle, adding to its growing body of verified decision procedures.

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

@@

To deal with this problem, we introduce a light-weight type system as displayed in \autoref{fig:types}. 
The type of a set term in this system is just a natural number which we call level.
Intuitively speaking, the level indicates that the corresponding term \lstinline!t! in Isabelle/HOL has type
\[
  \texttt{'a}\ \underbrace{\texttt{set}\ \ldots\ \texttt{set}}_{\texttt{l}\ \text{times}}
\]
for some \lstinline!'a!.
Note that the constructor $\emptyset$ now receives an additional argument that indicates the level of each instance of $\emptyset$.

@@

Ultimately, we define \lstinline!$\Gamma$ $\vdash$ $\psi$ $\equiv$ $\forall$a $\in$ atoms $\phi$. $\Gamma$ $\vdash$ a! in order to type formulae.

>>>>

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a quantifier-free fragment of set theory.
  We formalise its syntax and semantics as well as a sound and complete tableau calculus for it.
  We also provide an executable specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  Furthermore, we extend the calculus with a light-weight type system that paves the way for an integration of the procedure into Isabelle/HOL.
  \keywords{Decision procedures \and Tableau \and Proof assistants \and Set theory}
\end{abstract}

\section{Introduction}
In Isabelle/HOL, there are specialised procedures for dealing with e.g.\ natural numbers, linear arithmetic, and metric spaces.
Some of these procedures have been verified in Isabelle/HOL such as a procedure for Presburger arithmetic~\cite{presburger} that was later extended to mixed real-integer arithmetic~\cite{arithmetic}.
This procedure, though, uses reflection to work on goals in Isabelle/HOL, which, during execution, either sacrifices speed by going through the simplifier or requires trusting the code generator.
More recently, \citeauthor{orders}~\cite{orders} presented a verified decision procedure for orders that produces certificates. 
This approach offers efficient execution by using generated code as well as soundness because the certificates are replayed through Isabelle's inference kernel.

@@

Since \MLSS{} admits a tableau calculus, the generation of certificates will be straightforward.
Like with the aforementioned order solver, this paves the way towards an integration of the decision procedure into Isabelle, adding to its growing body of verified decision procedures.

\subsection{Contributions}
We present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
We also formalise a refinement to an executable specification from which Isabelle can generate code.
The formalisation initially follows the paper but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
\end{itemize}

@@

All in all, the formalisation amounts to over 6000 lines of theory.
It is available online~\cite{formalisation}.

@@

To deal with this problem, we formalise a light-weight type system as displayed in \autoref{fig:types}. 
The type of a set term in this system is just a natural number which we call level.
Intuitively speaking, the level indicates that the corresponding term \lstinline!t! in Isabelle/HOL has type
\[
  \texttt{'a}\ \underbrace{\texttt{set}\ \ldots\ \texttt{set}}_{\texttt{l}\ \text{times}}
\]
for some \lstinline!'a!.
Note that the constructor $\emptyset$ now receives an additional argument that indicates the level of each instance of $\emptyset$.

@@

Ultimately, we define \lstinline!$\Gamma$ $\vdash$ $\phi$ $\equiv$ $\forall$a $\in$ atoms $\phi$. $\Gamma$ $\vdash$ a! in order to type formulae.
